---
title: "12: Generalized Linear Models (Linear Regression)"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2020"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## Objectives
2. Apply special cases of the GLM (linear regression) to real datasets
3. Interpret and report the results of linear regressions in publication-style formats
3. Apply model selection methods to choose model formulations

## Set up
```{r, message = FALSE}
getwd()
library(tidyverse)
options(scipen = 4) # this is an option to use where it shows the number of digits in a result rather than giving expoential notation (so here, 4 digits but can set to anything)

PeterPaul.chem.nutrients <- read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")

# Set theme
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")
theme_set(mytheme)
```

## Linear Regression
The linear regression, like the t-test and ANOVA, is a special case of the **generalized linear model** (GLM). A linear regression is comprised of a continuous response variable, plus a combination of 1+ continuous response variables (plus the error term). The deterministic portion of the equation describes the response variable as lying on a straight line, with an intercept and a slope term. The equation is thus a typical algebraic expression: 
$$ y = \alpha + \beta*x + \epsilon $$

The goal for the linear regression is to find a **line of best fit**, which is the line drawn through the bivariate space that minimizes the total distance of points from the line. This is also called a "least squares" regression. The remainder of the variance not explained by the model is called the **residual error.** 

The linear regression will test the null hypotheses that

1. The intercept (alpha) is equal to zero.
2. The slope (beta) is equal to zero

Whether or not we care about the result of each of these tested hypotheses will depend on our research question. Sometimes, the test for the intercept will be of interest, and sometimes it will not.

Important components of the linear regression are the correlation and the R-squared value. The **correlation** is a number between -1 and 1, describing the relationship between the variables. Correlations close to -1 represent strong negative correlations, correlations close to zero represent weak correlations, and correlations close to 1 represent strong positive correlations. The **R-squared value** is the correlation squared, becoming a number between 0 and 1. The R-squared value describes the percent of variance accounted for by the explanatory variables. 

## Simple Linear Regression
For the NTL-LTER dataset, can we predict irradiance (light level) from depth?
```{r}
irradiance.regression <- lm(PeterPaul.chem.nutrients$irradianceWater ~ PeterPaul.chem.nutrients$depth)
# another way to format the lm function
irradiance.regression <- lm(data = PeterPaul.chem.nutrients, irradianceWater ~ depth)
summary(irradiance.regression)

# Correlation
cor.test(PeterPaul.chem.nutrients$irradianceWater, PeterPaul.chem.nutrients$depth)
# This gives R value, square to get R2 and should match summary output above
```
Question: How would you report the results of this test (overall findings and report of statistical output)?

> Some options:
> At greater depth, irradiance decreases (linear regression, R2=0.31, df=15,449, p<0.0001).
> Depth accounts for 31% of variance in lake irradiance (linear regression, R2=0.31, df=15,449, p<0.0001).
> For each 1 m increase in depth, irradiance decreases by 95 units (find units) (linear regression, R2=0.31, df=15,449, p<0.0001).
> Irradiance decreases significantly with decreasing depth (linear regression, R2=0.31, df=15,449, p<0.0001).

So, we see there is a significant negative correlation between irradiance and depth (lower light levels at greater depths), and that this model explains about 31 % of the total variance in irradiance. Let's visualize this relationship and the model itself. 

An exploratory option to visualize the model fit is to use the function `plot`. This function will return four graphs, which are intended only for checking the fit of the model and not for communicating results. The plots that are returned are: 

1. **Residuals vs. Fitted.** The value predicted by the line of best fit is the fitted value, and the residual is the distance of that actual value from the predicted value. By definition, there will be a balance of positive and negative residuals. Watch for drastic asymmetry from side to side or a marked departure from zero for the red line - these are signs of a poor model fit.

2. **Normal Q-Q.** The points should fall close to the 1:1 line. We often see departures from 1:1 at the high and low ends of the dataset, which could be outliers. 

3. **Scale-Location.** Similar to the residuals vs. fitted graph, this will graph the squared standardized residuals by the fitted values. 

4. **Residuals vs. Leverage.** This graph will display potential outliers. The values that fall outside the dashed red lines (Cook's distance) are outliers for the model. Watch for drastic departures of the solid red line from horizontal - this is a sign of a poor model fit.

```{r, fig.height = 3, fig.width = 4}
par(mfrow = c(2,2), mar=c(1,1,1,1)) # mar allows us to print when we knit the document
# This is just a verification of the assumptions of our test, never should end up in final report
# Residuals v. fitted plot: want to see horizontal line and no clustering of points in one location (same with scaled location plot), look for outliers too
plot(irradiance.regression)
par(mfrow = c(1,1))
# This sets R back to giving one plot at a time instead of in quadrants
```

The option best suited for communicating findings is to plot the explanatory and response variables as a scatterplot. 

```{r, fig.height = 3, fig.width = 4}
# Plot the regression
irradiancebydepth <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = depth, y = irradianceWater)) +
  ylim(0, 2000) +
  geom_point() 
print(irradiancebydepth) 
# One big outlier, remove by changing y axis limit
# Exponential decay relationship, might want to think about log transforming some of our variables to better fit a linear model, might also want to think about whether to take out outlier
# If log transforming values ofo 0 will give an error, in this case since so few, just remove them
```

Given the distribution of irradiance values, we don't have a linear relationship between x and y in this case. Let's try log-transforming the irradiance values.

```{r, fig.height = 3, fig.width = 4}
# Removing data pooints where irradiance is 0, so filtering to maintain values that are not equal to 0 and that are lower than 5000
PeterPaul.chem.nutrients <- filter(PeterPaul.chem.nutrients, 
                                   irradianceWater != 0 & irradianceWater < 5000)
# Log transforming
# Always go with adjusted R2 since this accounts for sample size and df
# No longer have very problematic departures from linear relationship, no extremely concerning outliers here
irradiance.regression2 <- lm(data = PeterPaul.chem.nutrients, log(irradianceWater) ~ depth)
summary(irradiance.regression2)

par(mfrow = c(2,2), mar=c(1,1,1,1))
plot(irradiance.regression2)
par(mfrow = c(1,1))

# Add a line and standard error for the linear regression
irradiancebydepth2 <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = depth, y = irradianceWater)) +
  # Geom_smooth is great for getting a line of best fit for 2 variables
  geom_smooth(method = "lm") +
  scale_y_log10() +
  # Kateri prefers logging the scale of the axis rather than the variable, easier to interpret she thinks
  geom_point() 
print(irradiancebydepth2) 

# SE can also be removed
irradiancebydepth2 <- 
    ggplot(PeterPaul.chem.nutrients, aes(x = depth, y = irradianceWater)) +
    geom_point(pch = 1, color = "dodgerblue") +
  # pch changes the point type, another option: can set alpha = 0.5 to change transparency
    scale_y_log10() +
  labs(x = "Depth (m)", y = "Irradiance (units)") +
    geom_smooth(method = 'lm', se = FALSE, color = "black")
print(irradiancebydepth2)

# Make the graph attractive

```

## Non-parametric equivalent: Spearman's Rho
As with the t-test and ANOVA, there is a nonparametric variant to the linear regression. The **Spearman's rho** test has the advantage of not depending on the normal distribution, but this test is not as robust as the linear regression.

``` {r}
cor.test(PeterPaul.chem.nutrients$irradianceWater, PeterPaul.chem.nutrients$depth, 
         method = "spearman", exact = FALSE)
```

## Multiple Regression
It is possible, and often useful, to consider multiple continuous explanatory variables at a time in a linear regression. For example, total phosphorus concentration in Paul Lake (the unfertilized lake) could be dependent on depth and dissolved oxygen concentration: 

``` {r, fig.height = 3, fig.width = 4}
TPregression <- lm(data = subset(PeterPaul.chem.nutrients, lakename == "Paul Lake"), 
                   tp_ug ~ depth + dissolvedOxygen)
summary(TPregression)
# Don't have to change data frame, but this subsets the dataset to only observations corresponding to Paul Lake

TPplot <- ggplot(subset(PeterPaul.chem.nutrients, lakename == "Paul Lake"), 
                 aes(x = dissolvedOxygen, y = tp_ug, color = depth)) +
  geom_point() +
  xlim(0, 20)
print(TPplot)

par(mfrow = c(2,2), mar=c(1,1,1,1))
plot(TPregression)
par(mfrow = c(1,1))

```

## Correlation Plots
We can also make exploratory plots of several continuous data points to determine possible relationships, as well as covariance among explanatory variables. 

```{r, fig.height = 3, fig.width = 4}
#install.packages("corrplot")
library(corrplot)
PeterPaulnutrients <- 
  PeterPaul.chem.nutrients %>%
  select(tn_ug:po4) %>%
  na.omit()
# This omits any row in the dataset that has even 1 NA
PeterPaulCorr <- cor(PeterPaulnutrients) # cor makes correlation matrix
corrplot(PeterPaulCorr, method = "ellipse")
corrplot.mixed(PeterPaulCorr, upper = "ellipse") # this gets rid of duplicates in matrix
```

## AIC to select variables

However, it is possible to over-parameterize a linear model. Adding additional explanatory variables takes away degrees of freedom, and if explanatory variables co-vary the interpretation can become overly complicated. Remember, an ideal statistical model balances simplicity and explanatory power! To help with this tradeoff, we can use the **Akaike's Information Criterion (AIC)** to compute a stepwise regression that either adds explanatory variables from the bottom up or removes explanatory variables from a full set of suggested options. The smaller the AIC value, the better. 

Let's say we want to know which explanatory variables will allow us to best predict total phosphorus concentrations. Potential explanatory variables from the dataset could include depth, dissolved oxygen, temperature, PAR, total N concentration, and phosphate concentration.

```{r}
# Smaller the AIC value the better, more parsimonious model
Paul.naomit <- PeterPaul.chem.nutrients %>%
  filter(lakename == "Paul Lake") %>%
  na.omit()
# Paul lake is the control lake while Peter lake is where nutrients dumped in to test for eutrophication

TPAIC <- lm(data = Paul.naomit, tp_ug ~ depth + dissolvedOxygen + 
              temperature_C + tn_ug + po4)
step(TPAIC) # Create new model using what AIC test tells you has lowest AIC
TPmodel <- lm(data = Paul.naomit, tp_ug ~ dissolvedOxygen + temperature_C + tn_ug)
summary(TPmodel)
```
> Result: Dissolved oxygen and temperature are significant predictors of total phosphorous (R2=0.258, df=3 and 109, p<0.001).